---
title: "LLM Demonstration: Eval Median"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LLM Demonstration: Eval Median}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>")
library(flowengineR)
library(jsonlite)

demo_dir <- system.file("llm_demonstrations", "2025-08-29_eval_median", package = "flowengineR")
```

# ðŸ“ Overview

This vignette documents a complete LLM-assisted engine build in *flowengineR*.  
The example shows how the evaluation engine `eval_median` was generated, registered, and executed.  
All inputs, transcripts, outputs, and provenance are archived under:

```{r}
demo_dir
list.files(demo_dir, recursive = TRUE)
```

# ðŸ“‚ Prompt & Inputs

```{r}
prompt_file <- file.path(demo_dir, "prompt", "llm_prompt_eval.R")
cat(readLines(prompt_file), sep = "\n")
```

# ðŸ’¬ Transcript

The full LLM conversation is included as a Quarto file:

```{r}
file.path(demo_dir, "conversation", "transcript.qmd")
```

# âš™ï¸ Generated Engine

```{r}
engine_file <- file.path(demo_dir, "outputs", "engine_eval_median.R")
cat(readLines(engine_file), sep = "\n")
```

# ðŸ§¾ Provenance

Console log and build call:

```{r}
cat(readLines(file.path(demo_dir, "provenance", "console_build.txt")), sep = "\n")
```

Session info and workflow results:

```{r}
cat(readLines(file.path(demo_dir, "provenance", "sessionInfo.txt"))[1:15], sep = "\n")
```

# ðŸ“Š Workflow Results

```{r}
res_file <- file.path(demo_dir, "provenance", "workflow_results.rds")
if (file.exists(res_file)) {
  results <- readRDS(res_file)
  str(results, max.level = 2)
}
```

# ðŸ”’ Integrity

Manifest with MD5 checksums:

```{r}
manifest <- fromJSON(file.path(demo_dir, "manifest.json"))
manifest$files
```
