---
title: "Getting Started with flowengineR"
subtitle: "First steps and basic workflow introduction"
author: "mwiller"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with flowengineR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> **Summary**  
> This vignette provides a **getting started guide** for `flowengineR`.  
> It covers installation, basic concepts of engines and workflows, and demonstrates a minimal example to help users set up their first experiments quickly.

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  library(flowengineR)
)
```

## 🚀 Introduction

The core strength of `flowengineR` is its **fully modular architecture**: any engine—whether used for data preprocessing, core algorithm execution, evaluation, or reporting—can be replaced or extended without changing the surrounding workflow.

This means you can:
- Test different preprocessing or postprocessing strategies with the same core algorithm
- Swap the algorithm engine (e.g., switch from a statistical model to a simulation) without touching the reporting logic
- Extend the workflow to entirely new domains, such as robustness testing, explainability, or compliance validation

> ⚡ In short: You only need to care about the "connection points"—not the entire workflow structure.
> 
> All default engines are just examples. The real power lies in defining your own logic using the framework’s unified interface.

`flowengineR` is a general-purpose workflow framework for algorithmic experimentation and analysis. One of its primary use cases is **fairness-aware modeling**, for which several ready-to-use engines are included.

This vignette introduces the basic workflow by walking through a minimal example.

You will learn how to:
- Define variables and model inputs
- Configure and run a training engine
- Apply a post-processing method
- Evaluate model performance and fairness

---

## 🎮 Interactive Start Menu

To make the first steps with `flowengineR` easier, the package provides an **interactive start menu**.  
This menu helps you explore the most important entry points, including running the example workflow, browsing vignettes, or inspecting the structure of control objects.

```r
# Launch the start menu
flowengineR_start()
```

When you run the function, you will see a menu in your console:

```
Where would you like to start?
1: ▶ Run example workflow
2: ☰ Open 'Index' vignette for overview of all vignettes
3: ℹ Open 'Getting Started' vignette
4: ⏺ List available engines
5: ❓ Help for run_workflow()
6: ★ Show structure of control-object
7: ℹ Open 'How to build custom Engines' vignette
8: ℹ Open 'How to use LLM-Engine-Builder' vignette
9: ✖ Exit
```

Choose an option by typing its number. For example:

- **1** runs the example workflow and assigns the results to `results_example`.  
- **4** lists all registered engines and assigns them to `engine_list`.  
- **6** shows the structure of the control object (`control_example`).  

This interactive helper is a great way to explore `flowengineR` without memorizing function names from the start.

---

## 🔧 Step 1: Define Variables

```{r}
vars = controller_vars(
  feature_vars = c(
    "profession.Employee", "profession.Selfemployed", "profession.Unemployed",
    "marital_status.Divorced", "marital_status.Married", "marital_status.Single",
    "housing_status.Own", "housing_status.Rent", "housing_status.WithParents",
    "region.Rural", "region.Suburban", "region.Urban",
    "employment_length", "credit_history_length", "number_prior_loans",
    "income", "loan_amount", "credit_score", "loan_to_income"
  ),
  protected_vars = c(
    "gender.Male", "gender.Female",
    "age"
  ),
  target_var = "default",
  protected_vars_binary = c(
    "gender.Male", "gender.Female",
    "age_group.<30", "age_group.30-50", "age_group.50+"
  )
)
```

---

## 🧠 Step 2: Define the Control Object

```{r}
control <- list(
  settings = list(
    log = list(
      log_show = TRUE,
      log_level = "info"
    )
  ),
  engine_select = list(
    train = "train_lm",
    postprocessing = "postprocessing_fairness_genresidual",
    eval = c("eval_statisticalparity", "eval_mse")
    ),
  params = list(
    train = controller_training(
      formula = default ~ income + loan_amount,
      norm_data = TRUE
    ),
    postprocessing = controller_postprocessing(),
    evaluation = controller_evaluation()
  )
)
```

A detailed introduction to the control object can be found [here](how_to_build_controlobject.html).

---

## ▶️ Step 3: Run the Workflow

```{r}
results <- run_workflow(control)
```

This returns a list of standardized results:
- model and predictions
- adjusted predictions (if fairness post-processing is used)
- evaluation metrics

---

## 📊 Step 4: Inspect Results

```{r}
results$metrics  # all computed metrics
results$model    # trained model object
```

---

## 🧪 Logging and Debugging

`flowengineR` includes a built-in logging system to improve transparency and simplify debugging throughout the entire workflow.

Logging is **enabled and configured via the `settings` section** of the `control` object:

```{r}
settings = list(
  log = TRUE,
  log_level = "debug"
)
```

Each internal function—especially within engines—can issue structured log messages through the centralized `log_msg()` helper. Messages are color-coded in the console and categorized by severity:

| Level   | Description                        | Console Color |
|---------|------------------------------------|---------------|
| `debug` | Detailed diagnostics               | Grey          |
| `info`  | Standard process messages          | Blue          |
| `warn`  | Recoverable issues and warnings    | Yellow        |
| `error` | Critical errors (can abort)        | Red           |

#### Example

```
#> [INFO] Starting run_workflow
#> [DEBUG] Calling training engine: train_lm
#> [WARN] Postprocessing skipped: no predictions found
#> [ERROR] Protected attributes missing (aborted)
```

This allows flexible diagnostics and helps users trace issues without manually inspecting intermediate objects.

---

## 🧩 Overview of Modular Engine Types

One of the core principles of **flowengineR** is modularity through interchangeable components called *engines*—these include engines for training, splitting, execution, evaluation, preprocessing, and reporting.

| Engine Type       | Workflow Position  | Description                                                                 | Input (via Wrapper)                                        | Output (via Initializer)                   | Example Engines                      |
| ----------------- | ------------------ | --------------------------------------------------------------------------- | ---------------------------------------------------------- | ------------------------------------------ | ------------------------------------ |
| [`split_*`](detail_engines_split.html)         | Dataset Splitting  | Splits the full dataset into training and test subsets                      | `data`, `target_var`, `seed`, `params`                     | `splits`, `split_type`, `seed`             | `split_random`, `split_stratified`   |
| [`execution_*`](detail_engines_execution.html)     | Execution Strategy | Executes the workflow logic across all splits (sequentially or in parallel) | `control`, `split_output`, `params`                        | `workflow_results`, `execution_type`       | `execution_sequential`               |
| [`preprocessing_*`](detail_engines_preprocessing.html) | Pre-Processing     | Modifies input data before algorithm execution                              | `data`, `params`, `protected_attributes`, `target_var`     | `preprocessed_data`, `method`              | `preprocessing_fairness_resampling`  |
| [`train_*`](detail_engines_train.html)         | Core Algorithm     | Fits a model or runs an algorithm on training data                          | `formula`, `data`, `norm_data`, `params`                   | `model`, `model_type`, `predictions`       | `train_lm`, `train_gbm`              |
| [`inprocessing_*`](detail_engines_inprocessing.html)  | In-Processing      | Alters internal training logic (e.g., constraints, weighting)               | `driver_train`, `data`, `params`, `protected_attributes`   | `adjusted_model`, `predictions`            | `inpro(..)_adversialdebiasing`       |
| [`postprocessing_*`](detail_engines_postprocessing.html)| Post-Processing    | Modifies predictions after model execution                                  | `predictions`, `params`, `protected_attributes`            | `adjusted_predictions`, `method`           | `postprocessing_fairness_genresidual`|
| [`eval_*`](detail_engines_evaluation.html)          | Evaluation         | Calculates performance or bias metrics                                      | `predictions`, `actuals`, `params`, `protected_attributes` | `metrics`, `eval_type`                     | `eval_statisticalparity`, `eval_mse` |
| [`reportelement_*`](detail_engines_reportelements.html) | Reporting Element  | Creates tables or plots from results                                        | `workflow_results`, `split_output`, `alias`, `params`      | `report_object`, `report_type`             | `reportelement_table_splitmetrics`   |
| [`report_*`](detail_engines_report.html)        | Report Composition | Assembles report elements into a structured document                        | `reportelements`, `alias_report`, `params`                 | `report`, `sections`, `compatible_formats` | `report_modelsummary`                |
| [`publish_*`](detail_engines_publish.html)       | Output Publishing  | Exports objects or results to external targets                              | `object`, `file_path`, `alias_publish`, `params`           | `path`, `success`, `type`                  | `publish_excel_basis`                |


---

## 🔍 Listing Available Engines

To get an overview of all currently registered engines, use the helper function `list_registered_engines()`:

```r
# Show all registered engines, grouped by type
list_registered_engines()
```

This will output all available engines, grouped by engine type such as `train`, `split`, `execution`, `evaluation`, `preprocessing`, and `postprocessing`.

You can also filter the output to a specific engine type:

```r
# Show only execution engines
list_registered_engines(type = "execution")
```

This function is especially useful if:
- you have registered your own custom engines and want to verify their integration,
- you are unsure about the correct naming of available engines,
- you want to introspect the internal engine setup during debugging or development.

It is a key utility for working transparently and flexibly within the flowengineR framework.

---

## ➕ Next Steps

- [Engine Development Guide](how_to_build_custom_engine.html)
- [Metalevel Flowchart](metalevel_flowchart.html)
